Exploring Patterns in Clinical and Non-Clinical Listeria monocytogenes Isolates 

### 2025-05-02 code/analysis-code/statistical-analysis.qmd
additions

# Plot
seasonalpreds <- ggplot(combined_data, aes(x = month, y = clinical_count, color = source)) +
  geom_line() +
  labs(title = "Actual and Forecasted Clinical Isolate Counts",
       y = "Clinical Count",
       x = "Month") +
  scale_color_manual(values = c("Actual" = "black", "Forecast" = "blue")) +
  theme_minimal()
print(seasonalpreds)
ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/seasonalpreds.png", plot = seasonalpreds, width = 8, height = 6, dpi = 300)
```

>>>>

# Plot
seasonalpreds <- ggplot(combined_months, aes(x = month, y = clinical_count, color = source)) +
  geom_line() +
  labs(title = "Actual and Forecasted Clinical Isolate Counts",
       y = "Clinical Count",
       x = "Month") +
  scale_color_manual(values = c("Actual" = "black", "Forecast" = "blue")) +
  theme_minimal()
print(seasonalpreds)
ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/seasonalpreds.png", plot = seasonalpreds, width = 8, height = 6, dpi = 300)
```

### 2025-05-02 code/eda-code/eda.qmd
additions

Load the data.

@@

Showing a bit of code to produce and save a summary table.

```{r}
summary_df = skimr::skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("results", "tables", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

# Data exploration through figures

@@

@@

locations <- ggplot(
  mydata %>% 
    filter(food == "Yes", Location != "USA"),
  aes(x = Location)
) + 
  geom_bar(fill = "steelblue") + 
  labs(
    title = "Count of Food-Isolated Listeria Cases Across Locations",
    x = "Location",
    y = "Count of Yes"
  ) + 
  theme_minimal() + 
  theme(axis.text.x = element_text(size = 6, angle = 45, hjust = 1))
print(locations)
ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/locations.png", plot = locations, width = 8, height = 6, dpi = 300)
```

@@

mydata %>%
  dplyr::filter(clinical == "Yes") %>%  
  count(`Isolation source`) %>%
  ggplot(aes(x = `Isolation source`, y = n)) +
  geom_boxplot(fill = "darkblue") +
  labs(title = "Count of Clinical 'Yes' Across Isolation Sources", 
       x = "Isolation Source", 
       y = "Count of Yes") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 4, angle = 45, hjust = 1))

@@

```{r}
mydata %>% 
  dplyr::filter(clinical == "Yes") %>%
  count(`Isolation source`) %>%
  top_n(10, n) %>%  # Keep only the top 10 sources
  ggplot(aes(x = reorder(`Isolation source`, -n), y = n)) +
  geom_bar(stat = "identity", fill = "darkblue") +
  labs(title = "Top 10 Clinical 'Yes' Isolation Sources", 
       x = "Isolation Source", 
       y = "Count of Yes") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1))

@@

```{r}
mydata %>% 
  dplyr::filter(food == "Yes") %>%
  count(`Isolation source`) %>%
  top_n(10, n) %>%  # Keep only the top 10 sources
  ggplot(aes(x = reorder(`Isolation source`, -n), y = n)) +
  geom_bar(stat = "identity", fill = "darkblue") +
  labs(title = "Top 10 Food 'Yes' Isolation Sources", 
       x = "Isolation Source", 
       y = "Count of Yes") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1))
```
#We see here that the most food isolates for listeria monocytogenes are found in Ready To Eat products, that is a term used by Food Safety and Inspection Service to describe any food products that do not require cooking or sterilization to safely consume.

@@

# Step 1: Prepare the data — monthly counts
monthly_counts <- mydata %>%
  mutate(month = floor_date(`Create date`, "month")) %>%
  group_by(month) %>%
  summarize(
    food_count = sum(food == "Yes"),
    clinical_count = sum(clinical == "Yes")
  ) %>%
  ungroup()

# Step 2: Convert to long format for plotting
monthly_long <- monthly_counts %>%
  pivot_longer(cols = c(food_count, clinical_count),
               names_to = "source_type",
               values_to = "count")

# Step 3: Plot
ggplot(monthly_long, aes(x = month, y = count, color = source_type)) +
  geom_line(size = 1.2) +
  labs(
    title = "Monthly Salmonella Isolate Counts by Source",
    x = "Month",
    y = "Isolate Count",
    color = "Source Type"
  ) +
  scale_color_manual(values = c("food_count" = "forestgreen", "clinical_count" = "firebrick")) +
  theme_minimal()
```

```{r}
Isolatecounts <- ggplot(monthly_long, aes(x = month, y = count, fill = source_type)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Monthly Salmonella Isolate Counts by Source (Stacked Bars)",
    x = "Month",
    y = "Isolate Count",
    fill = "Source Type"
  ) +
  scale_fill_manual(values = c("food_count" = "forestgreen", "clinical_count" = "firebrick")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/isolatecounts.png", plot = Isolatecounts, width = 8, height = 6, dpi = 300)
print(Isolatecounts)
```

>>>>

#Load the data.

@@

@@

#Table 1: counts across the 15 year period with range and mean

```{r}
library(kableExtra)

# Create new year and month columns
mydata <- mydata %>%
  mutate(
    year = year(`Create date`), 
    month = month(`Create date`, label = TRUE, abbr = TRUE)
  )

# Count records
counts <- mydata %>%
  group_by(year, month, isolation_source) %>%
  summarize(count = n(), .groups = "drop")

# Summarize by year and source
yearly_totals <- counts %>%
  group_by(isolation_source, year) %>%
  summarize(
    total = sum(count),
    highest_month = month[which.max(count)],
    lowest_month = month[which.min(count)],
    .groups = "drop"
  )

# Pivot for final table
table_data <- yearly_totals %>%
  pivot_wider(
    names_from = isolation_source,
    values_from = c(total, highest_month, lowest_month),
    names_sep = "_"
  ) %>%
  arrange(year)

# Create the styled table
table_output <- table_data %>%
  kbl(
    caption = "Yearly Isolate Summary by Source",
    col.names = c(
      "Year", 
      "Total Clinical", "Total Environmental", "Total Food", 
      "Highest Month (Clinical)", "Highest Month (Environmental)", "Highest Month (Food)", 
      "Lowest Month (Clinical)", "Lowest Month (Environmental)", "Lowest Month (Food)"
    )
  ) %>%
  kable_styling(
    full_width = FALSE, 
    position = "center", 
    bootstrap_options = c("striped", "hover", "condensed")
  )
```

```{r}
# Define the file path
summarytable_file <- here::here("results", "tables", "summarytable.rds")

# Save the styled table as an RDS file
saveRDS(table_output, file = summarytable_file)

@@

locations <- ggplot(
  mydata %>% 
    filter(food == "Yes", Location != "USA"),
  aes(x = Location)
) + 
  geom_bar(fill = "steelblue") + 
  labs(
    title = "Count of Food-Isolated Listeria Cases Across Locations",
    x = "Location",
    y = "Count of Isolates"
  ) + 
  theme_minimal() + 
  theme(axis.text.x = element_text(size = 6, angle = 45, hjust = 1))
print(locations)
ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/locations.png", plot = locations, width = 8, height = 6, dpi = 300)
```

@@

mydata %>%
  dplyr::filter(clinical == "Yes") %>%  
  count(`Isolation source`) %>%
  ggplot(aes(x = `Isolation source`, y = n)) +
  geom_boxplot(fill = "darkblue") +
  labs(title = "Count of Clinical Isolates Across Isolation Sources", 
       x = "Isolation Source", 
       y = "Count of isolates") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 4, angle = 45, hjust = 1))

@@

```{r}
mydata %>% 
  dplyr::filter(clinical == "Yes") %>%
  count(`Isolation source`) %>%
  top_n(10, n) %>%  # Keep only the top 10 sources
  ggplot(aes(x = reorder(`Isolation source`, -n), y = n)) +
  geom_bar(stat = "identity", fill = "darkblue") +
  labs(title = "Top 10 Clinical 'Yes' Isolation Sources", 
       x = "Isolation Source", 
       y = "Count of isolates") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1))

@@

```{r}
mydata %>% 
  dplyr::filter(food == "Yes") %>%
  count(`Isolation source`) %>%
  top_n(10, n) %>%  # Keep only the top 10 sources
  ggplot(aes(x = reorder(`Isolation source`, -n), y = n)) +
  geom_bar(stat = "identity", fill = "darkblue") +
  labs(title = "Top 10 Food Isolation Sources", 
       x = "Isolation Source", 
       y = "Count of Yes") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1))
```
#We see here that the most food isolates for listeria monocytogenes are found in Ready To Eat products, that is a term used by Food Safety and Inspection Service to describe any food products that do not require cooking or sterilization to safely consume.

@@

#Prepare the data into monthly counts
monthly_counts <- mydata %>%
  mutate(month = floor_date(`Create date`, "month")) %>%
  group_by(month) %>%
  summarize(
    food_count = sum(food == "Yes"),
    clinical_count = sum(clinical == "Yes"),
    environmental_count = sum(isolation_source =="environmental")
  ) %>%
  ungroup()

#Convert to long format for plotting
monthly_long <- monthly_counts %>%
  pivot_longer(cols = c(food_count, clinical_count, environmental_count),
               names_to = "source_type",
               values_to = "count")

#Plot monthly counts
ggplot(monthly_long, aes(x = month, y = count, color = source_type)) +
  geom_line(size = 1.2) +
  labs(
    title = "Monthly Salmonella Isolate Counts by Source",
    x = "Month",
    y = "Isolate Count",
    color = "Source Type"
  ) +
  scale_color_manual(values = c("environmental_count" = "forestgreen", "food_count" = "firebrick", "clinical_count"= "darkblue")) +
  theme_minimal()
```

```{r}
Isolatecounts <- ggplot(monthly_long, aes(x = month, y = count, fill = source_type)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Monthly Salmonella Isolate Counts by Source (Stacked Bars)",
    x = "Month",
    y = "Isolate Count",
    fill = "Source Type"
  ) +
  scale_fill_manual(values = c("environmental_count" = "forestgreen", "food_count" = "firebrick", "clinical_count"= "darkblue")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/isolatecounts.png", plot = Isolatecounts, width = 8, height = 6, dpi = 300)
print(Isolatecounts)
```

### 2025-05-02 code/processing-code/processingfile-v1.qmd
additions

```{r}
library(readxl) 
library(dplyr) 
library(tidyr)
library(skimr) 
library(here) 
```

# Data loading

Note that for functions that come from specific packages (instead of base R), I often specify both package and function like so:
package::function() that's not required one could just call the function specifying the package makes it clearer where the function "lives",
but it adds typing. You can do it either way.

```{r}
# path to data
# note the use of the here() package and not absolute paths
data_location <- here::here("data","raw-data","lmonoisolates.xls")
rawdata <- readxl::read_excel(data_location)
```

@@

```{r}
summary(rawdata)
```

Several ways of looking at the data

```{r}
dplyr::glimpse(rawdata)
summary(rawdata)
head(rawdata)
skimr::skim(rawdata)
```

```{r}
unique(rawdata$BioSample)
```
```{r}
unique(rawdata$Location)
```
```{r}
unique(rawdata$`Create date`)
```

@@

After inspecting the data,I have dedcided to remove some variables that I will not use. This will be serovar, assembly, biosample for now. I want to try and make a map so i will keep in isolate names and snp distances and what not.

```{r}
lmono2 <- rawdata %>% 
  select(-c(Serovar, Assembly, BioSample))
```
Next I want to reformat the dates so that it removes the timestamp. I will just be keeping the year, month and day
```{r}
lmono2$`Create date` <- as.Date(lmono2$`Create date`)
print(lmono2$`Create date`)
```
Next I will see the range of dates and select a 15 year time frame.
```{r}
lmono3 <- lmono2 %>%
  filter(`Create date` >= as.Date("2010-01-01") & `Create date` <= as.Date("2024-12-31"))
```
Finally I will check for any missing entries
```{r}
any(is.na(lmono3))
lmono3[!complete.cases(lmono3), ]  # Shows me what rows have missing data

@@

@@

# Save data

Finally, we save the clean data as RDS file. I suggest you save your processed and cleaned data as RDS or RDA/Rdata files. 
This preserves coding like factors, characters, numeric, etc.  If you save as CSV, that information would get lost.
However, CSV is better for sharing with others since it's plain text. If you do CSV, you might want to write down somewhere what each variable is.

See here for some suggestions on how to store your processed data:
http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata

```{r}
save_data_location <- here::here("data","processed-data","processeddata.rds")
saveRDS(processeddata, file = save_data_location)
```

>>>>

```{r}
library(readxl) 
library(dplyr) 
library(tidyr)
library(skimr) 
library(here) 
library(lubridate)
```

# Data loading
```{r}
# path to data
# note the use of the here() package and not absolute paths
data_location <- here::here("data","raw-data","lmonoisolates.xls")
rawdata <- readxl::read_excel(data_location)
```

@@

```{r}
summary(rawdata)
dplyr::glimpse(rawdata)
summary(rawdata)
head(rawdata)
skimr::skim(rawdata)
```

@@

```{r}
lmono2 <- rawdata %>% 
  select(-c(Serovar, Assembly, BioSample))
```
Next I want to reformat the dates so that it removes the timestamp. I will just be keeping the year, month and day
```{r}
lmono2$`Create date` <- as.Date(lmono2$`Create date`)
print(lmono2$`Create date`)
```

```{r}
lmono2 <- lmono2 %>%
  mutate(
    year = year(`Create date`),
    month = month(`Create date`, label = TRUE, abbr = TRUE)
  ) %>%
  filter(year != 2011)

```
Next I will see the range of dates and select a 15 year time frame.
```{r}
lmono3 <- lmono2 %>%
  filter(`Create date` >= as.Date("2010-01-01") & `Create date` <= as.Date("2024-12-31"))
```
Finally I will check for any missing entries
```{r}
any(is.na(lmono3))
lmono3[!complete.cases(lmono3), ]  # Shows me what rows have missing data

@@

#Creating an additional variable called "Isolation_source" This is different from the `Isolation Source` variable in that it will include food and not just list it under environmental/other
```{r}
lmono_final <- lmono_final %>%
  mutate(
    clinical = as.character(clinical),
    food = as.character(food),
    isolation_source = case_when(
      clinical == "Yes" ~ "clinical",
      food == "Yes" ~ "food",
      TRUE ~ "environmental"
    )
  )

```

@@

# Save data 
```{r}
save_data_location <- here::here("data","processed-data","processeddata.rds")
saveRDS(processeddata, file = save_data_location)
```

### 2025-05-02 products/manuscript/Manuscript.qmd
updated readme files
additions

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

@@

For temporal analysis was conducted using poisson regression. All machine learning models were conducted using the tidymodels workflow. Specific packages for poisson modeling were parsnip and poissonreg.

@@

@@

The first goal of this analysis is to find any seasonal patterns in *Listeria* infections and *Listeria* contaminated food items.

>>>>

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
library(htmltools)
```

@@

Temporal analysis was first visualized using

conducted using poisson regression. All machine learning models were conducted using the tidymodels workflow. Specific packages for poisson modeling were parsnip and poissonreg.

@@

```{r}
table_output
```

@@

The first goal of this analysis is to find any seasonal patterns in *Listeria* infections and *Listeria* contaminated food items. In figure 5 we can see that cases of listeriosis typically dip around the 100th day of the year which is early April, and spikes near the 200th to 300th days of the year which are mid July to late October. The prediction model forecasts low counts of Listeriosis from the end of 2023 to earlt 2024
